---
title: "Week 6"
author: "Daniel Gonzalez-Suarez"
date: "3/3/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---
# Introduction
Here we will explore how to create basic plots in R. To acknowledge my forebears, much of the material on this week's class is based on a course in R by Dr. Brett Favaro.

Read the file WhitefishLS.csv into R
 These data are derived from Derek Ogle's FSA data resources:
  http://derekogle.com/fishR/data/data-html/WhitefishLS.html

>>> Whitefish example, Variables are:
year: Year of catch.
state: State of catch (MI, MN, or WI).
catch: Catch in lbs.
value: Value of catch in dollars.

# Install required packages
```{r, message=FALSE}
# If you don't yet have tidyverse run:
# install.packages("tidyverse")
library(tidyverse)

# We will require this package to handle some dates in this dataset
# install.packages("lubridate")
library(lubridate)

```

#Load data: Remember to check if it loaded & if data types are correct
```{r}
whitefish <- read.csv("./data/WhitefishLS.csv")

# ------------------------------------------
#STEP 1: Did it load correctly?
  head(whitefish) 

# ------------------------------------------
#STEP 2: Are the data types right?
  sapply(whitefish, class) 

#appears fine, except we have a date-type variable here.

# We will use the lubridate::date_decimal
# It tells R this is a date not a number
whitefish$year <- date_decimal(whitefish$year)

sapply(whitefish, class) 
# We will explore some other functions of lubridate later

# ------------------------------------------
# STEP 3: Check for missing or impossible values

range(whitefish$year) #It now lines up with http://derekogle.com/fishR/data/data-html/WhitefishLS.html

range(whitefish$catch) #nothing jumps out as wrong. I guess it's a big fishery
plot(whitefish$catch)


range(whitefish$value) #nothing impossible
plot(whitefish$value)

# ------------------------------------------
# STEP 4: Factor levels

levels(whitefish$state)
# All seems to be OK!

# Let's move on to plotting!

```

# Plot One Variable - Base
```{r}
plot(whitefish$catch ~ whitefish$state)

# One variable, X = Discrete (states)
# Y can only be count (catch)

table(whitefish$state)

#Here 'table' creates a contingency table of the counts of each factor in a specific variable, in this case the number of times a state is mentioned
?table

barplot(table(whitefish$state))

plot(table(whitefish$state))

# PLOT ONE VARIABLE
# Continuous X

plot(whitefish$catch) 
# No, this does not work

hist(whitefish$catch) 
# Yes, this looks much better

?hist

# We can chose different breaks or bins of data
hist(whitefish$catch, breaks=10)
hist(whitefish$catch, breaks=20)
hist(whitefish$catch, breaks=30)

# Or assign them mannually
hist(whitefish$catch, breaks=c(0,500000,1000000,1500000, 2000000))
hist(whitefish$catch, breaks=c(0,100000,200000,300000,400000,500000,1000000,1500000,2000000))

# Many calculations are conducted by the hist function prior to plotting
# and you can extract those values
histinfo <- hist(whitefish$catch)

histinfo

# Making a point about data density... let's plot mean by state
# Remember some of dplyr functions
temp <- whitefish %>%
  group_by(state) %>%
  summarise(avg = mean(catch))

plot(temp$avg ~ temp$state)
# It is nice to have average values, but they can mislead interpretation
# of results

# Perhaps it is better to have range of values on this one
# This can be easily achieved with boxplot
boxplot(whitefish$catch ~ whitefish$state)
# When you plot a Discrete X and continuous Y variable plot will return 
# a boxplot by default
plot(whitefish$catch ~ whitefish$state)

```

# Plot One Variable - ggplot
```{r}
# Formulation and syntaxis is different

ggplot(data = whitefish,
       aes(state))  +
  geom_bar()

#ggplot(data = DATA,
#       aes(WHAT YOU WANT TO PLOT)) +
#  GRAPHICAL ELEMENTS, ADDED ONE BY ONE LIKE LAYERS

a <- ggplot(data = whitefish,
       aes(state)) +
  geom_bar()

print(a) #this is necessary when in the R interface
a #this one also works in Rstudio

# Demonstrate how things can be layered
a <- ggplot(data = whitefish,
            aes(state))

a #Here we just called the function and the discrete variable but have not assigned what will be plotted

# ONE VARIABLE CONTINUOUS

a <- ggplot(data=whitefish,
            aes(catch))
# OK, we want to plot catch values

a + geom_area(stat="bin") #warning just tells you that default 'bin=30'
a + geom_density()
a + geom_dotplot()#warning just tells you that default 'bin=30'
a + geom_freqpoly()
a + geom_histogram() #this one is like base 'hist()'


#Let's follow the warning suggestion and explore different bin sizes
a + geom_histogram(bins=10)
a + geom_histogram(bins=20)
a + geom_histogram(bins=30)
a + geom_histogram(bins=40) 

a + geom_histogram(bins = 10) + coord_flip()
#'coord_flip' flips the axis order

```

# Plot Two Variable - Base
```{r}
plot(value ~ catch, data=whitefish)

plot(catch ~ value, data=whitefish)

# Which is the correct one? Which story do you want to tell? 
# Which one makes more sense?
# Remember 'response' vs 'predictive' variables

plot(catch ~ year, data=whitefish)
# We can add lines representing linear model fits
abline(lm(catch~year, data=whitefish))

plot(year ~ catch, data=whitefish)
abline(lm(catch~year, data=whitefish))
# This won't work, it does not make sense variables in this order


```

# Plot Two Variable - ggplot
```{r}
# Continuous X, continuous Y

a <- ggplot(data=whitefish, aes(x = year, y = catch))
# The above creates an object where we store the 'base plot'
a + geom_point() # This adds points as a scatterplot

a <- ggplot(data=whitefish, aes(x = year, y = catch)) +
  geom_point() +
  geom_smooth(method="lm", level=0.95) # Then add a linear model fit

a # Call to see how it looks

?geom_smooth() # This one creates the linear model fit

ggplot(data=whitefish, aes(x = year, y = catch)) +
  geom_point() +
  geom_smooth(span=100) 

# Loess smoothing is a process by which many statistical softwares do smoothing

# -------------------------------
# Discrete X, continuous Y

# Recall in base plot this will return a boxplot
plot(catch ~ state, data=whitefish)

# In ggplot 'geom_boxplot()' would achieve this
ggplot(data=whitefish, aes(x = state, y = catch)) +
  geom_boxplot()

# Adding datapoints may reveal some pattern not easy to see in a whiskerplot
ggplot(data=whitefish, aes(x = state, y = catch)) +
  geom_jitter() + # This prevents data points from overlapping
  geom_boxplot(alpha=0.7)

a <- ggplot(data=whitefish, aes(x = state, y = catch))

a + geom_boxplot()


# -------------------------------
# Discrete X vs. discrete Y

whitefish <- whitefish %>%
  mutate(BigState = ifelse(state=="MI", "Y", "N"))


ggplot(data=whitefish, aes(x=state, y = BigState)) +
  geom_count()


```
