---
title: "Week 6, exercises"
author: "Daniel Gonzalez-Suarez"
date: "3/4/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Introduction
Using "iris" database, we'll explore how to make plots in Base R and ggplot.

# Install required packages
Don't forget to load the packages you're going to use for this project.
```{r, message=FALSE}
library(tidyverse)

```

#Verification of data.
We'll have to check ir everything's okay before starting to plot.
```{r}
# ------------------------------------------
#STEP 1: Did it load correctly?
head(iris)
tail(iris)
# ------------------------------------------
#STEP 2: Are the data types right?
str(iris)
sapply(iris, class) 
# ------------------------------------------
# STEP 3: Check for missing or impossible Petal.Lengths #
range(iris$Petal.Length)
range(iris$Sepal.Length)
range(iris$Sepal.Width)
range(iris$Petal.Width)
plot(iris)

# ------------------------------------------
# Step 4: Check factor levels
levels(iris$Species)

```

# Plotting one variable with Base R
```{r}
# One variable, X = Discrete (Species)
# Y can only be count of cases
table(iris$Species)
barplot(table(iris$Species))

# Continuous X (Plot only one of the four in the dataset)
plot(iris$Species, iris$Sepal.Length)
plot(iris$Species, iris$Sepal.Width)
plot(iris$Species, iris$Petal.Length)
plot(iris$Species, iris$Petal.Width)
     
  
# Using hist() 
hist(iris$Sepal.Length, breaks = 4)
hist(iris$Sepal.Length, breaks = 6)

hist(iris$Sepal.Width, breaks = 3)
hist(iris$Sepal.Width, breaks = 7)

```

# Let's use ggplot
```{r}
# One variable, X = Discrete (Species)
# Y can only be count of cases

ggplot(iris, aes(Species)) +
  geom_bar()

# Create an object to store the plot and add more layers latter 
plot_sp <- ggplot(iris, aes(Species)) +
  geom_bar() 

# You can flip the axis order
plot_sp + coord_flip()

```

# Plot Two Variable - Base

```{r}
# Continuous X, continuous Y


# Remember these questions when preparing these plots:
# Which is the correct one? Which story do you want to tell? Which one makes more sense? Depends on your question.
plot(Petal.Length ~ Sepal.Length, data=iris)
plot(Sepal.Length ~ Petal.Length, data=iris)

# Try adding lines representing linear model fits
plot(Sepal.Length ~ Petal.Length, data=iris)
abline(lm(Sepal.Length~Petal.Length, data=iris))

plot(Petal.Length ~ Sepal.Length, data=iris)
abline(lm(Petal.Length ~ Sepal.Length, data=iris))


# -------------------------------
# Discrete X, continuous Y

# Make a plot that most accurately represent the whole range of a continuous (Y) by species (X)
plot(Petal.Length ~ Species, data=iris)
plot(Petal.Width ~ Species, data=iris)
plot(Sepal.Length ~ Species, data=iris)
plot(Sepal.Width ~ Species, data=iris)

```

# Plot Two Variable - ggplot
```{r}
# Continuous X, continuous Y
# Explore different plots interchanging the 'predictive' and 'response' variables among the four available
plotSLPW <- ggplot(iris, aes(Sepal.Length, Petal.Width)) + 
  geom_point()

plotSLSW <- ggplot(iris, aes(Sepal.Length, Sepal.Width)) + 
  geom_point()

plotPLSW <- ggplot(iris, aes(Petal.Length, Sepal.Width)) + 
  geom_point()

plotPLPW <- ggplot(iris, aes(Petal.Length, Petal.Width)) + 
  geom_point()

plotSLPW 
plotSLSW 
plotPLSW 
plotPLPW 

# Try adding lines representing different trend lines to help you explore patterns
# Add a linear model fit to your plots

plotSLPW + geom_smooth(method = "lm")
plotSLSW + geom_smooth(method = "lm")
plotPLSW + geom_smooth(method = "lm")
plotPLPW + geom_smooth(method = "lm")

# Add a 'Loess smoothing' fit to your plots
plotSLPW + geom_smooth(method = "loess")
plotSLSW + geom_smooth(method = "loess")
plotPLSW + geom_smooth(method = "loess")
plotPLPW + geom_smooth(method = "loess")

# -------------------------------
# Discrete X, continuous Y

# Make a plot that most accurately represent the whole range of a continuous (Y) by species (X)
plot_sp <- ggplot(iris, aes(Species, Sepal.Length)) +
  geom_boxplot()


# Try adding datapoints to this plot to reveal patterns not easy to see in a whiskerplot
plot_sp + 
  geom_boxplot(alpha=0.69) +
  geom_jitter()


```


